---
title: "Invasives_percentages_all"
format: html
---

### Overview

Workflow to collate percentage invasion and RLE assessment of Criterion D per Vegtype for each invasive species data set. This workflow produces a summary table that is taken up in the RLE compiled assessment.

It draws on the primary outputs from workflows for NIAPS (INV_terr/Invasives_niaps.qmd), CFR IAT (Invasives_rebelo_wc.qmd), MAPWAPs (Invasives_mapwap.qmd), Umgeni_Tugela (Invasives_umngeni.qmd).

```{r}
library(tidyverse)
options(scipen =999)

niaps <- read.csv("outputs/data_for_rle_niaps.csv")   %>% 
  select(T_MAPCODE, niaps_ext70, D3_niaps, B1i_niaps, B2i_niaps) %>%
  rename(prpinv_niaps = niaps_ext70 )

bioscape <- read.csv("outputs/data_for_rle_rebelo_invwc.csv") %>% 
  select(T_MAPCODE, prop_inv, D3_cfr_iat, B1i_cfr_iat, B2i_cfr_iat)%>%
  rename(prpinv_cfs_iat = prop_inv )

umgeni <- read.csv("outputs/data_for_rle_um.csv")%>% 
  select(T_MAPCODE, prop_inv, D1_um, B1i_um, B2i_um)%>%
  rename(prpinv_umgeni = prop_inv )

mapwaps <- read.csv("outputs/data_for_rle_mapwaps.csv")%>% 
  select(T_MAPCODE, prop_inv, D1_mapwaps, B1i_mapwaps, B2i_mapwaps)%>%
  rename(prpinv_mapwaps = prop_inv )

```

```{r}
#combine invasive results and select max category and max perc invaded

inv_comb <- niaps %>%
  left_join(bioscape, by = "T_MAPCODE") %>%
  left_join(mapwaps, by = "T_MAPCODE") %>%
  left_join(umgeni, by = "T_MAPCODE") %>%
  mutate(across(matches("^[DB]"), ~ recode(.x,
                                           "CR" = 4L,
                                           "EN" = 3L,
                                           "VU" = 2L,
                                           "LC" = 1L,
                                           .default = 0L,
                                           .missing = 0L))) %>%
  rowwise() %>%
  mutate(
    prpinv_max = max(c_across(matches("^prpinv")), na.rm = TRUE),
    D3_inv    = max(c_across(matches("^D")), na.rm = TRUE),
    B1i_inv   = max(c_across(matches("^B1")), na.rm = TRUE),
    B2i_inv   = max(c_across(matches("^B2")), na.rm = TRUE)
  ) %>%
  ungroup()


  # write results
write.csv(inv_comb, file = "outputs/data_for_rle_inv_comb.csv")              
   

```
