---
title: "Combined invasive alien plant layer for the NBA 2025"
bibliography: references.bib
lang: en-GB
---

### Overview of analysis:

This workflow combines invasive alien plant data from various sources into a single raster coverage. This will be used as an additional input layer in updated Red List of Threatened Plant and Amphibian assessments, and terrestrial Ecosystem Protection Level Assessments for the National Biodiversity Assessment 2025.

#### Data sources:

1.  Map of woody invasive alien plant taxa within the Cape Floristic Region. This classification was generated using Sentinel-2 satellite imagery at a 10 m resolution using a Random Forest machine learning classifier. The data is unprocessed and contain certainty information in Band 2 (values of 0-100) ([@rebelo]). The data were prepossessed in ARCGIS PRO: Band one and two were combined such that only pixels with greater than 50% certainty were used, and all invasive species classes were recoded to value = 8. Non invaded areas were coded value = 0. areas outside of AOI were coded value = 99. The data were resampled from 10m to 20m pixels and snapped to the land cover 2022 7 class grid. Projection set to Albers Equal Area Conic CM: 25, STDP1: -24, STDP2: -33, WGS84.

2.  Map of key invasive alien plant taxa within Mzimvubu, Tugela, Sabie-Crocodile and Luvuvhu catchment areas collected as part of the MAPWAPS program ([@skosana], [@cogill], @skosanaa\], [@cogilla] ). This classification was generated using Sentinel-2 satellite imagery at a 10 m resolution using a Gradient Tree Boost (GTB) classification. The data were pre processed in ARCGIS PRO, where an attribute \[mapwaps\] was added to each data set. Where LUCL = Alien\* the value 8 was assigned, otherwise the value 0. This allows for determination of invaded areas and other areas within the area of interest. The shapefiles for Luvuhvu and and Sabie-Crocodile were projected from UTM36S to UTM35S, and then all four were merged into a single shapefile mapwaps.shp. this was then rasterized (using mapwaps attribute) and snapped to the national land cover grid 2022. Projection set to Albers Equal Area Conic CM: 25, STDP1: -24, STDP2: -33, WGS84.

3.  National Invasive Alien Plant Survey (NIAPS) ([@kotzé2025]) estimated the extent of the most-widespread & abundant, terrestrial invasive alien plant taxa (approx. 32 species) in South Africa. Data were downloaded from an ARCPRO package available [here](https://dffeportal.environment.gov.za/portal/home/item.html?id=17de13c509ef4d3caf279d84e77312c3). The data excludes invasions in transformed areas (i.e. cultivated, built-up & planted forest). Each raster has pixel values (0-100) that are percentage of area invaded divided by condensed area invaded for 32 Invasive plant taxa organised into 13 rasters. Values of 100 represent 100% invasion (effectively 100% canopy cover of the specific invasive species)(see [@marais2004] for explanation of the concept of "condensed area"). The data were prepossessed in ARCGIS PRO: All rasters were combined using a sum, pixels with values \> 50 were recoded to value = 8, those then than 50% were coded value = 0. The data were resampled from decimal degrees to 20m pixels and snapped to the land cover 2022 7 class grid. Projection set to Albers Equal Area Conic CM: 25, STDP1: -24, STDP2: -33, WGS84.

4.  National Land Cover 7 class 2022.

#### Analysis and outputs: 

The three invasives rasters were combined such that any pixel of value = 8 (invaded) was retained and all othe rpixels were assigned value =0 (not invaded) to produce a inv_comb.tif raster.

This combined invasion raster was combined with the 7 class national land cover such that if land cover was = 1 (natural) and the combined invasion layer = 8 (invaded) then the output pixel value was = 8. If not then the land cover value was assigned. This results in a 8 class national land cover : 1 = natural, 2 = secondary natural(old fields), 3 = artificial water bodies, 4 = buyilt up (infrastructure), 5 = croplands and orchards, 6 = mines and mine dumps, 7 = plantation forestry, 8 = invasive alien plant (high density). Ths produced a lc2022_inv.tif raster.

```{r}
#load packages  
library(terra) 
library(tidyverse)
library(sf)
library(PROJ)
```

```{r}
# load niaps data - all species combined and using only pixels with > 50% condensed area (high canopy cover of iaps) 
niaps <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/Kotze_National_2024/data/niaps2.tif") 

#load Bioscape CFR data - all species combined and using only pixels with certainty > 50%
inv_cfr <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/Rebelo Invasives WC/inv_cert2.tif")  

#load MAPWAPS data - all species combined 
mapwaps <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/MAPWAPS/mapwaps2.tif")  
# extra step required - resample as mapwaps didnt have same extent - dont know why as in ARCPRO I snapped it to the LC2022. 
mapwaps <- resample(mapwaps, niaps, method = "near" )

# Create a single-layer mask where any raster has value = 8
has_8 <- sum(niaps == 8, inv_cfr == 8, mapwaps == 8) > 0  

# Apply the ifel() operation
inv_comb2 <- ifel(has_8, 8, 0)

# write to disk
writeRaster(inv_comb2, "C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/inv_comb.tif", datatype = "INT1U", gdal = c("COMPRESS=LZW"), overwrite = TRUE)

# load LC 2022
lc <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif")

# combine the rasters such that if lc = 1 or lc = 2 AND inv = 8 then value in new raster = 8, else use value of lc. This creates a 8 class land cover 1=nat, 2=secnat, 3=artwater, 4=built, 5=crop, 6=mine, 7=plant, 8=invaded

lc2022_inv <-  ifel(inv_comb2 == 8 & lc == 1, 8, lc)

writeRaster(lc2022_inv, "C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/lc2022_inv.tif", datatype = "INT1U", gdal = c("COMPRESS=LZW"), overwrite = TRUE)


```
