---
title: "RLE terrestial 2025 - D3 assessment using MAPWAPS Alien Tree data"
bibliography: references.bib
lang: en-GB
---

### Overview of analysis:

Exploring invasive alein plant data produced by the MAPWAPS project ([@cogill; @cogilla; @skosana; @skosanaa]) for use in terrestrial Red List of Ecosystems - a component of the National Biodiversity Assessment 2025.

The extent and severity of invasive alien plant invasions can be used as a confirmation of "ongoing decline" in IUCN RLE 1.1 Criteria B1 and B2 assessments (restricted distribution), and in Criterion D assessment of functional declines (biotic disruption). Since the invasions by alien plant species were reported more than 50 years ago ([@macdonald1984]), Criterion D3 was applied (disruption compared to a historic base line 1750). The approach followed was to cross tabulate South Africa's national vegetation map (version 2024) (SANBI, 2025) and the modified national land cover 2022 (SANBI, 2024) and alien plant species occurrence data from the MAPWAPS project ([@skosana], [@cogill], [@skosanaa] , [@cogilla] ) .

The severity of biotic disruption in pixels identified as invaded is estimated to be \>80%. Non natural areas (croplands, old fields, built up areas , mines, dams and plantations) were masked out of the analysis, and only natural remaining vegetation patches were assessed as to their extent of invasion within the area of interest.

```{r}
#load packages  
library(terra) 
library(tidyverse)
library(sf)
library(PROJ)
```

### Data sources & import:

1.  South African National Land Cover data set for 2022 (prepared by the National Department of Forestry, Fisheries and the Environment) were adjusted to have a common raster resolution (20m). The data were reclassified in ARCGIS PRO into seven classes: 1 = Natural; 2 = Secondary Natural, 3 = Artificial water bodies, 4 = Built up, 5 = Croplands, 6 = Mines, 7 = Plantation (SANBI pers com).

2.  National Vegetation Map 2024 version, vector data (ESRI file geodatabase) January 2025 curated by SANBI ([@Dayaram2019])

3.  Map of key invasive alien plant taxa within Mzimvubu, Tugela, Sabie-Crocodile and Luvuvhu catchment areas collected as part of the MAPWAPS program ([@skosana], [@cogill], [@skosanaa] , [@cogilla] ). This classification was generated using Sentinel-2 satellite imagery at a 10 m resolution using a Gradient Tree Boost (GTB) classification. The data were pre processed in ARCGIS PRO, where an attribute \[mapwap\] was added to each dataset. Where LUCL = Alien\* the value 8 was assigned, otherwise the value 0. This allows for determination of invaded areas and other areas within the area of interest. The shapefiles for Luvuhvu and and Sabie-Crocodile were projected from UTM36S to UTM35S, and then all four were merged into a sibgle shapefile mapwaps.shp. this was then rasterized (using mapwap attribute) and snapped to the national land cover grid 2022 (including reprojection from UTM to Albers Equal Area to match national land cover and vegetation data).

```{r}
#add invasive data : supplied by Coghill and Skosana (for use in RLE only)
mapwap <- rast("C:/Users/skownoa/Dropbox/NBAwork/RLEworking/Invasives/MAPWAPS/mapwaps2.tif")

#Add land cover data  (note all LC products use  WGS84 Albers Equal Area, central meridian = 25; std parallel1 = -24; std parallel2 = -33). Use Tif version as this is processed faster by R terra.
lc2022 <- rast("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/nlc2022_7class.tif")
#lc2022 <- resample(lc2022, inv_raw, method = "near")

# add vegetation data
veg <- vect("C:/Users/skownoa/Dropbox/NBAwork/NLC_change/NLC_Change.gdb", "NVM2024Final_IEM5_12_07012025")  #ensure it uses proj =  wgs84 aea cm 25 -24 -33

# to help differentiate masked areas (LES and SWA) from NON TERRESTRIAL areas add "NOTVEG" to T_MAPCODE where T_NAME = "Non-Terrestrial*"
veg$T_MAPCODE <- ifelse(veg$T_Name == "Non-terrestrial (estuarine habitat)" | veg$T_Name == "Non-terrestrial (Aquatic)", "NOTVEG", veg$T_MAPCODE) 
vegrast <- rasterize(veg,lc2022,"T_MAPCODE")

# ensure extents match 
mapwap <- resample(mapwap, lc2022, method = "near")

# load EOO and AOO for Criterion B assessment (ongoing decline)
eoo <- read_csv("C:/Rdata/RLE_terr/outputs/EOO_AOO_from_redlistr.csv")
```

### Spatial Analysis

Cross tabulate the land cover, vegetation and invasive data to produce a value matrix - then convert this to a tibble (tidy table). Add extent (km2) based on pixel counts (pixels are 10mx10m). Add vegetation information (Biome, Name and full extent) to the table.

```{r}
#Crosstabulate invasives, land cover and vegegation data 
#Convert raw output values to a tibble
mapwap_values <- crosstab(c(lc2022,vegrast,mapwap), useNA=TRUE) #ensure NA areas included
mapwap_values <- as_tibble(mapwap_values) #save raw crosstab outputs
write.csv(mapwap_values, file = "outputs/mapwap_values.csv") 

#--------------------------------------------------------------------
#--------------------------------------------------------------------
# Start here: load values to avoid rerunning time consuming cross tab above

mapwap_values <- read_csv("outputs/mapwap_values.csv", show_col_types = FALSE)

#Rename columns and add km2 (cell size is 10x10m), and make adjustments to values as required
mapwap_tb <- mapwap_values %>%
  mutate(km2 = (n*100)/1000000) %>% # add extent 10x10m pixels
  mutate(IAT = ifelse(IAT == "NaN", 0, IAT))%>% # make NaN = 0 (allows as.integer later)
  rename(LC = nlc2022_7class) %>%
  filter(n !=0, T_MAPCODE != "NOTVEG") %>%
  mutate(LC = as.integer(LC), IAT = as.integer(IAT), IAT_cert = as.integer(IAT_cert)) #IMPORTANT to make integers othewise ifelse does not work correctly below!

# # write output to csv for excel use
write.csv(inv_lc_veg_tb, file = "outputs/inv_wc_lc_veg_tb.csv") 

```

### Summary statistics

The extent of IAT invasion in the remaining natural extent of each ecosystem type is then expressed as a proportion. This proportion together with an estimation of the Relative Severity of biotic disruption from the invasion allows for the applicaiton of Criterion D of the Red List of Ecosystems. The pure pixel apporach suggests that IAT pixels have a high relative severity of \>= 90%. If the invasions are thought to have occured more than 50years ago then Criterion D3 is applied, if the majority of the invasion is though to have occurred within the last 50years then Criterion D1 can be applied.

|                              |               |               |               |
|------------------------------|---------------|---------------|---------------|
| Criterion D1                 | \>=80% extent | \>=50% extent | \>=30% extent |
| **\>=80% relative severity** | **CR**        | **EN**        | **VU**        |
| \>=50% relative severity     | EN            | VU            |               |
| \>=30% relative severity     | VU            |               |               |

|                              |               |               |               |
|------------------------------|---------------|---------------|---------------|
| Criterion D3                 | \>=90% extent | \>=70% extent | \>=50% extent |
| **\>=90% relative severity** | **CR**        | **EN**        | **VU**        |
| \>=70% relative severity     | EN            | VU            |               |
| \>=50% relative severity     | VU            |               |               |

```{r}
options(scipen = 999)
# Calculate the the extent of IAT per vegetation unit, but only where : landcover is natural (LC = 1) and where IAT certainty  is greater than 50% (IAT_cert>2) and invasives have been recored (IAT =1). To ensure focus on AOI use IAT_cert != 0 this means outside of AOI. But note IAT = 0 simply means no invasives detected.

vegrem_extent_in_aoi <- inv_lc_veg_tb %>%
  filter(LC == 1 , IAT_cert != 0) %>% # only natural extent in the aoi. 
  group_by(T_MAPCODE) %>%
  summarise(sumrem_aoi = sum(km2))
  
veg_extent_in_aoi <- inv_lc_veg_tb %>%
  filter(IAT_cert != 0) %>% # full extent in aoi
  group_by(T_MAPCODE) %>%
  summarise(sum_aoi = sum(km2))
 
# identify invaded areas with high certainty estimates, then summarise and work out prop of natural remaining vegetation that is invaded. This script requires that cript for niaps (Invasives_niaps.qmd) has already been run to produce the table "vegrem_extent_in_sa"

inv_veg_sum <- inv_lc_veg_tb %>%
  mutate(inv_deg = ifelse(IAT_cert > 5 & # could adjust IAT_cert up to 6 (>50% cert)
                           !is.na(IAT) &
                           IAT != 0 & # count only invaded areas
                           LC == 1, 1, 0)) %>% # count only invasion in natural veg
       group_by(T_MAPCODE, inv_deg) %>%
  summarise(sum_IAT = sum(km2)) %>%
  filter(inv_deg  == 1, !is.na(T_MAPCODE), T_MAPCODE != "NOTVEG") %>%
  left_join(vegrem_extent_in_aoi, by = "T_MAPCODE") %>%
  left_join(veg_extent_in_aoi, by ="T_MAPCODE") %>%
  left_join(vegrem_extent_in_sa, by = "T_MAPCODE") %>% # from niaps workflow
  mutate(prop_rem_aoi = sumrem_aoi/sumrem_sa) %>% # check how much of type is in AOI
  mutate(prop_inv = sum_IAT/sumrem_sa) %>%
  mutate(D3_cfr_iat = case_when(prop_inv >= 0.9 ~ "CR",
                   prop_inv >= 0.7 ~ "EN",
                   prop_inv >= 0.5 ~ "VU",
                  TRUE ~ "LC")) %>%
  left_join(eoo, "T_MAPCODE") %>%
  mutate(B1i_cfr_iat = case_when(EOO <= 2000 & prop_inv >= 0.3 ~ "CR", 
                                  EOO <= 20000 & prop_inv >= 0.3 ~ "EN", 
                                  EOO <= 50000 & prop_inv >= 0.3 ~ "VU", 
                                  TRUE ~ "LC")) %>%
  mutate(B2i_cfr_iat = case_when(AOO <= 2 & prop_inv >= 0.3 ~ "CR", 
                                  AOO <= 20 & prop_inv >= 0.3 ~ "EN", 
                                  AOO <= 50 & prop_inv >= 0.3 ~ "VU", 
                                  TRUE ~ "LC"))

#  write output to csv to hand over to RLE assessment (in the next step the proportion of each ecosystem )
write.csv(inv_veg_sum, file = "outputs/data_for_rle_rebelo_invwc.csv") 

```
